The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-13 pkg/partykit/inst/guideliketest/draft/draft.Rnw
Kramer -> Kraemer
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2554 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1f450d9448bba5f693d63a378f8be08f3e18ba7c

Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
model scores) perform clearly better than the original GUIDE version (using residuals 
only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
of replications. Even though the advantage of CTree and MOB evens out in certain situations
such as for a high effect size (see, e.g., for $\xi=0$ and varying $\beta_0$), GUIDE is 
never ahead of them in any of the considered settings. Moreover, the positive effect of 
including all available scores can be observed particularly well in the top middle panel 
where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
GUIDE. The fact that residual-based methods fail to detect changes that do not effect the
expected value has already been elaborated in the literature of structural change
analysis \citep[see, e.g.,][]{Ploberger+Kramer:1992} and is also discussed in 
Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
to GUIDE+scores in case of a fixed slope parameter (left panels).

@@

Linear models as presented in this study have been selected as they allow for a simple illustration
of the investigated subjects. However, the obtained results also hold for a variety of other models
as the introduced building blocks can be applied in the same way for all model types.
Specific characteristics of linear models regarding structural changes have been investigated
in detail in previous literature.
\cite{Ploberger+Kramer:1992} have shown that residual-based tests can detect a change in the parameters
only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
the parameter vector $\bm{\beta}$ is shifted 
%to $\bm{\beta} + \bm{\Delta}$ 
by $\bm{\Delta} = (0, \Delta_1)^{\top}$.
For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
While this might be an extreme example similar scenarios can appear in many applications, see, e.g., 
Figure~2 in \cite{Loh+He+Man:2015} where an additional separation of the data in each node is
required in order to fit a reasonable model. This can be avoided by including model scores
allowing for the detection of changes not only in the location parameter but also in all other 
parameters, i.e., not only in the expected value but also in the variance.
In situations where shifts appear only in the intercept testing procedures purely based on residuals 
perform best. However, the cost in terms of power that comes with including all scores, in particular
with testing more than one parameter, is negligibly small.

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
and \citealp{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kramer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of of testing all parameters.

>>>>

Figure~\ref{fig:scores_stump} shows that overall CTree and MOB (both using full 
model scores) perform clearly better than the original GUIDE version (using residuals 
only) as they accomplish to select the true split variable $Z_1$ in a higher proportion
of replications. Even though the advantage of CTree and MOB evens out in certain situations
such as for a high effect size (see, e.g., for $\xi=0$ and varying $\beta_0$), GUIDE is 
never ahead of them in any of the considered settings. Moreover, the positive effect of 
including all available scores can be observed particularly well in the top middle panel 
where GUIDE+scores detects a shift in the slope parameter $\beta_1$ as opposed to 
GUIDE. The fact that residual-based methods fail to detect changes that do not effect the
expected value has already been elaborated in the literature of structural change
analysis \citep[see, e.g.,][]{Ploberger+Kraemer:1992} and is also discussed in 
Section~\ref{sec:discussion}. Moreover, the advantage of including full models scores 
clearly outweighs the cost involved which is displayed by the small gain of GUIDE compared 
to GUIDE+scores in case of a fixed slope parameter (left panels).

@@

Linear models as presented in this study have been selected as they allow for a simple illustration
of the investigated subjects. However, the obtained results also hold for a variety of other models
as the introduced building blocks can be applied in the same way for all model types.
Specific characteristics of linear models regarding structural changes have been investigated
in detail in previous literature.
\cite{Ploberger+Kraemer:1992} have shown that residual-based tests can detect a change in the parameters
only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor $(1,0)^{\top}$, hence, if
the parameter vector $\bm{\beta}$ is shifted 
%to $\bm{\beta} + \bm{\Delta}$ 
by $\bm{\Delta} = (0, \Delta_1)^{\top}$.
For that reason GUIDE breaks down when the slope parameter $\beta_1$ changes while the intercept
$\beta_0$ is fixed as seen in the middle panel of Figure~\ref{fig:scores_stump}.
While this might be an extreme example similar scenarios can appear in many applications, see, e.g., 
Figure~2 in \cite{Loh+He+Man:2015} where an additional separation of the data in each node is
required in order to fit a reasonable model. This can be avoided by including model scores
allowing for the detection of changes not only in the location parameter but also in all other 
parameters, i.e., not only in the expected value but also in the variance.
In situations where shifts appear only in the intercept testing procedures purely based on residuals 
perform best. However, the cost in terms of power that comes with including all scores, in particular
with testing more than one parameter, is negligibly small.

Also in more general models residual-based and score-based procedures are expected to perform equally 
well if all parameters are highly correlated. On the other hand, if parameters are orthogonal changes 
that are not connected to the mean might be missed when only considering residuals. This can be prevented 
by including scores which are a standard measurement of goodness of fit that can be easily employed for 
all model types.
Even though it also depends on the type of the applied residuals 
(see, e.g., \citealp{Loh+Zheng:2013}, \citealp{Chaudhuri+Loh:2002} 
and \citealp{Loh+He+Man:2015})
whether parameter changes can be detected successfully, a one-dimensional measurement of goodness of 
fit such as residuals provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of of testing all parameters.
